ODE and Diffusion Models of Tumor Growth

### 2024-03-15 paper.tex
8.5

        \draw[->, thick] (0, 0) -- (0.45, 0) node[midway, above] {$r_1$};
        \draw[->, thick] (0, 0) -- (0, 1) node[midway, right] {$r_2$};
        \draw[->, thick] (0, 0) -- (-1.07, -1.07) node[midway, above] {$r_3$};
    \end{tikzpicture}
    \caption{A cross-section of the sphere that represents the tumor. The necrotic core is situated between $0 < r < r_1$,
        the middle layer is situated between $r_1 < r < r_2$, and the healthy outer layer is located between $r_2 < r < r_3$.}
    \label{fig:cross-section}
\end{figure}

Typically, the few tumor cells can expand quite rapidly depending on several parameters, such as the cell types and nutrient levels of the body.
They can multiply in number to around a million in a growth period of days. The nodule (Figure \ref{fig:cross-section}) can expand from being microscopic
to having a radius of over 1 millimeter. Greenspan wished to apply this model to understand how $r_3$ - the outer radius of the nodule - changes
with respect to time by analyzing the steady state. To do this, several assumptions were adopted by the model:

\begin{enumerate}
    \item[(i)] The tumor is a sphere with perfect symmetry at all times, leaving the experiment to have 2 independent variables: time, $t$, and $r$, the radial distance.
    \item[(ii)] Cancer cells die when the nutrient levels, $\sigma(r, t)$ drop below a critical minimum of $\sigma_{min}.$
\end{enumerate}

>>>>

        \draw[->, thick] (0, 0) -- (0.45, 0) node[midway, above] {$r_1(t)$};
        \draw[->, thick] (0, 0) -- (0, 1) node[midway, left] {$r_2(t)$};
        \draw[->, thick] (0, 0) -- (-1.07, -1.07) node[midway, right] {$r_3(t)$};
    \end{tikzpicture}
    \caption{A cross-section of the sphere that represents the tumor. The necrotic core is situated between $0 < r < r_1$,
        the middle layer is situated between $r_1 < r < r_2$, and the healthy outer layer is located between $r_2 < r < r_3$.}
    \label{fig:cross-section}
\end{figure}

Typically, the few tumor cells can expand quite rapidly depending on several parameters, such as the cell types and nutrient levels of the body.
They can multiply in number to around a million in a growth period of days. The nodule (Figure \ref{fig:cross-section}) can expand from being microscopic
to having a radius of over 1 millimeter. Greenspan wished to apply this model to understand how $r_3(t)$ - the outer radius of the nodule - changes
with respect to time by analyzing the steady state. To do this, several assumptions were adopted by the model:

\begin{enumerate}
    \item[(i)] The tumor is a sphere with perfect symmetry at all times, leaving the experiment to have 2 independent variables: time, $t$, and $r$, the radial distance.
    \item[(ii)] Cancer cells die when the nutrient levels, $\sigma(r, t)$ drop below a critical minimum of $\sigma_{min}.$ Oxygen is the most essential nutirient and is modeled using the methods described in the paper by Hill (1928).
    \item[(iii)] Oxygen is used by living cells only and depends on the nutrient concentration.
\end{enumerate}

The model also assumes that living cancer cells maintain a compact solid mass and that any mass lost from the diffusion of permeable substances coming from
disintegrated necrotic cells is replaced by cells pushing towards the center of the compact mass. Because of this, we arrive at the conclusion in equilibrium that,
if the rate of cell production exceeds the losses from necrotic disintegration, then the tumor must be growing in size. The reverse of this statement also holds.

Recall from above that the middle layer of the tumor has some substance that inhibits the mitosis of cancer cells. We define $\beta(r, t)$ to be the concentration
of this substance in cubic centimeters.

\subsection{Compartments and Conservation of Mass}

We need a few additional approximations before establishing the model:

\begin{enumerate}
    \item[(i)] All living tumor cells are the same and do not fluctuate in size.
    \item[(ii)] Mitosis occurs instantaniously and all new cells are the same size as existing cells.
    \item[(iii)] The growth rate via mitosis is a distribution $S(\sigma, \beta)$ with $S(\sigma, \beta) = 0$ for $\beta > \beta_{max}$ or $\sigma < \sigma_{min}$.
    \item[(iv)] The core loses cell volume proportional to its volume with the proportionality constant defined as $3\lambda$.
\end{enumerate}

Now we can discuss the conservation of mass law:

\begin{equation}
    c_{total}(t) = c_0 + c_{prod}(t) - n(t) - n_{lost}(t)
\end{equation}
where

$c_{total}(t)$ is the total number of cells produced at time $t$,

$c_0$ is the inital population,

$c_{prod}(t)$ is the total amount of cells produced during the interval $[0, t)$,

$n(t)$ is the total volume of necrotic cells at $t$,

and $n_{lost}(t)$ is the volume lost to the necrotic core during the interval $[0, t)$.

Let $r_3(0)$ be the inital outer radius of the tumor. Recall that cell death occurs when the nutrient level $\sigma$ falls below
its minimum value $\sigma_{min}$, the edge of the inner radius, $r_1(t)$, must satisfy

\begin{equation}
    \sigma(r_1(t), t) = \sigma_{min}, t \geq 0
\end{equation}

Let $r_2(t)$ be the radius where the chemical inhibitor reaches the critical level. Then by definition

\begin{equation}
    \beta(r_2(t), t) = \beta_{max}
\end{equation}

We can now replace the formulas described above with mathematical expressions:

\begin{equation}
    c_{total}(t) = \frac{4\pi}{3}(r_3^3(t) - r_1^3(t))
\end{equation}

by subtracting the areas of the two spheres.

It is quite obvious to see that

\begin{equation}
    c_{init}(t) = \frac{4\pi}{3}r_3^3(0)
\end{equation}

and that

\begin{equation}
    n(t) = \frac{4\pi}{3}r_1^3(t)
\end{equation}

Using our assumption that the necrotic core loses volume at a constant rate, we can proudce the integral

\begin{equation}
    n_{lost}(t) = \frac{4\pi}{3} \int_{0}^{t}3\lambda r_1^3(t) dt
\end{equation}
